NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all

### 2022-11-23 main.tex
Updates from Overleaf

%\newpage

@@

\ref{DatabankStructure}
\ref{fig:permeationSI}

\subsection{Molecule and atom naming convention} \label{naming}
When analysing simulations, atoms and molecules needs to be often called by their names defined in the simulation trajectory. However, these names often vary between the force fields because the universal naming convention has not been defined for lipids. To enable automatic analyses over simulations with different atom and molecule names in the NMRlipids databank, we have defined unique naming conventions for molecules and atoms therein. Unique abbreviations used in the NMRlipids databank for each molecule are listed in table~\ref{tab:abbreviations} in the supplementary information. Atom names in a simulation are connected to unique atom names defined for each molecule using mapping files located at {\it /Scripts/BuildDatabank/mapping\_files} in the NMRlipids repository. 
%Mapping files are in yaml format where unique atom names are connected to force field specific names. 
Also residue names for each atom and whether they belong to headgroup, glycerol bakcbone, or acyl chain region are defined in the mapping files. For more details about mapping files, see~\url{https://nmrlipids.blogspot.com/2022/04/new-yaml-format-of-mapping-files.html}). In practise, 
%information about molecule and atom names are given, where 
mapping file names and force field specific molecule names corresponding the unique abbreviation (table~\ref{tab:abbreviations}) are given in the COMPOSITION dictionary in README.yaml files. 
%Upon addition of a new entry in the databank, the molecule names in the simulation corresponding the unique names and mapping files (available at \url{https://github.com/NMRLipids/Databank/tree/main/Scripts/BuildDatabank/mapping_files}) are defined in the dictionary.

\subsection{Adding data into the NMRlipids databank}
Anyone can add a new simulation into the NMRlipids databank by first creating an info.yaml file containing the manually entered information listed in table~\ref{tab:READMEkeys} and then adding this file to {\it /Scripts/BuildDatabank/info\_files} folder in the NMRlipids databank repository via pull requests. Pull requests are manually accepted and automatically extracted data for README.yaml files,  listed in table~\ref{tab:READMEkeys}, will be calculated using the {\it /Scripts/BuildDatabank/AddData.py} script. Current content of the NMRlipids databank is composed of the simulations found from Zenodo repository with an appropriate licence. Most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

@@

The quality of each C--H bond order parameter is estimated by calculating the probability of a simulated value to locate within the error bars of the experimental value. Because conformational ensembles of individual lipids are independent in a fluid lipid bilayer, 
$\frac{S_{\rm CH}-\mu}{s/\sqrt{n}}$ has a Student's t-distribution with $n-1$ degrees of freedom and the probability for a order parameter in simulation to locate within experimental error bars can be estimated from equation
\begin{equation}\label{probability}
  P = f \left( \frac{S_{\rm CH} - (S_{\rm exp}+\Delta S_{\rm exp})}{s/\sqrt{n}} \right) - f \left( \frac{S_{\rm CH} - (S_{\rm exp}-\Delta S_{\rm exp})}{s/\sqrt{n}} \right),
\end{equation}
where f(t) is the %first order 
Student's t-distribution, $\mu$ is the real mean of the order parameter, $n$ is the number of independent sample points for each C-H bond which equals the number of lipids in a simulation, $S_{\rm CH}$ is the sample mean from Eq.~\ref{OP}, $s$ is the variance of $S_{\rm CH}$ calculated over individual lipids, $S_{\rm exp}$ is the experimental value, and $\Delta S_{\rm exp}$ its error. The error of $\Delta S_{\rm exp} = 0.02$ is currently assumed for all experimental order parameters~\cite{ollila16}, yet more accurate may be available in the future~\cite{wurl22}. Because lipid bilayer simulations contain at least dozens of lipids, the Student's t-distribution could be safely approximated with a normal distribution. However, the normal distribution gives probability values that are below the numerical accuracy of computers when simulation values are far from experiments. To avoid such numerical instabilities, we use the first order Student's t-distribution having slightly higher probabilities for values far away from the mean.
%Because phospholipids sample their conformational ensemble within nanosecond timescale~\cite{ferreira15}, all simulations in the databank would be sufficiently long to sample the realistic conformational phase of individual lipids. 
On the other hand, some force fields exhibit too slow dynamics which leads to large error bars in order parameter values~\cite{antila21a}. Such large error bars widen the Student's t-distribution in Eq.~\ref{probability} thereby artificially increasing the probability to find the simulated value within experimental error bars. Therefore, the order parameters with simulation error bars above the experimental error 0.02 are not included in the quality evaluation.

@@

Also a quality measure based on differences in simulated and experimental form factors accross available q-range is defined in the SIMtoEXP program~\cite{kucerka10}. However, the lobe heigths in simulated form factors depend on the simulation box size as shown in Fig.~\ref{fig:sizedependence}. Therefore, the quality measure defined in SIMtoEXP  would also depend on the simulation box size. Nevertheless, locations of form factor minima are independent on simulation box size in Fig.~\ref{fig:sizedependence}. Here we use the location of the first form factor minima for the quality evaluation because automatic detection of the location of second minima is inaccurate for some experimental data due to fluctuations, such as for the POPE data in Figs.~\ref{fig:quality} D and E. 
%In order to define a quality measure which is independent on simulation box size, we measure the form factor quality based on the location of the first minima observed in experiments. 
The first minimum correlates well with the thickness of a membrane (Fig.~\ref{fig:quality} F), although the correlation of the second minima would be even stronger (Fig.~\ref{fig:QualityCorrelationsSI}). In practise, we first filter the fluctuations from the form factor data using Savitzky-Golay filter (window lenght 30 and polynomial order 1) and locate the first minima above 0.1\,$AA^2$ from both simulation and experimental data. The quality of a form factor is then defined as Eucledian distance between the minima in simulated and experimental form factors, $FF_q = |FF_{\rm{min}}^{\rm{sim}}-FF_{\rm{min}}^{\rm{exp}}|$.

@@

\subsection{Analysing anisotropic diffusion of water in membranes environment from the NMRlipids databank}

@@

Water permeability through membranes was calculated from equation $P=r/2c_w$, where $r$ is the rate of permeation events per time and area, and $c_w$=33.3679\,(nm)$^{-3}$ which is the concentration of water in bulk~\cite{venable19}. The number of permeation events in each trajectory was calculated using the code by Camilo et al.~\cite{camilo2022}, available at~\url{https://github.com/crobertocamilo/MD-permeation}. 
%This code was used within the loop that goes over all simulations in the NMRlipids databank and extracts the required information using the information in README.yaml files. 
The code that calculates permeabilities for all simulations in the NMRlipids databank is available at {\it /scripts/calcMD-PERMEATION.py} and the resulting permeabilities are stored at {\it /Data/MD-PERMEATION} in the repository containing all analyses specific for this publication at~\url{https://github.com/NMRLipids/DataBankManuscript/}. 
%While the order parameters, form factors, area per lipid and thickness are stored within the NMRlipids databank (\url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}), further analyses can be conventiently stored in separate repositories with the same folder structure based on hash identities of trajectory and topology files. For example, results from further analyses performed here are stored in folders at \url{https://github.com/NMRLipids/DataBankManuscript/tree/main/Data}. 
This repository is organized similarly to the NMRlipids databank repository, enabling the upcycling of also the analyzed data without overloading the main NMRlipids databank repository.

@@

\subsection{Calculation of lipid flip-flops}
Flip-flop rates were calculated using {\it AssignLeaflets} and {\it FlipFlop} tools from LiPyphilic package~\cite{LiPyphilic2021}. Headgroup atoms of each molecule defined in the mapping file were used to determine in which leaflet the molecule locates. The midplane cut-off defining the region between leaflets was 1\,nm and frame cut-off was 100. This means that if the headgroup of a molecule entered within the distance of 1\,nm from the bilayer midplane and was found from the opposing leaflet after 100 steps, this event was considered as a successfull flip-flop event. A code that finds flip-flop rates from all simulations in the databank is available at {\it scripts/FlipFlop.py} and the resulting flip-flop events at {\it Data/Flipflops/} in the repository at~\url{https://github.com/NMRLipids/DataBankManuscript/}.

>>>>

\newpage

@@

\subsection{Molecule and atom naming convention} \label{naming}
When analysing simulations, atoms and molecules needs to be often called by their names defined in the simulation trajectory. However, these names typically vary between force fields because the universal naming convention has not been defined for lipids. To enable automatic analyses over simulations with different atom and molecule names in the NMRlipids databank, we have defined unique naming conventions for molecules and atoms therein. Unique abbreviations used in the NMRlipids databank for each molecule are listed in table~\ref{tab:abbreviations} in the supplementary information. Atom names used in simulation trajectories are connected to unique atom names using mapping files that are defined in the NMRlipids project (\url{https://nmrlipids.blogspot.com/2022/04/new-yaml-format-of-mapping-files.html}). These files are located at {\it /Scripts/BuildDatabank/mapping\_files} in the NMRlipids databank repository. These files also define whether an atom belongs to headgroup, glycerol backbone, or acyl chain region in a lipid. In practise, force field specific molecule names and mapping files
names are given in the COMPOSITION dictionary in README.yaml files for each molecule in each simulation in the NMRlipids databank.

\subsection{Adding data into the NMRlipids databank}
The NMRlipids databank is open for additions of simulation data by anyone. The first step is to create an info.yaml file containing the information that needs to be manually entered as listed in table~\ref{tab:READMEkeys}. This file can be then added into {\it /Scripts/BuildDatabank/info\_files} folder in the NMRlipids databank repository via pull requests. After the pull request is manually accepted, the rest of the information for the README.yaml file, listed in table~\ref{tab:READMEkeys}, will be automatically extracted using the {\it /Scripts/BuildDatabank/AddData.py} script. Currently the NMRlipids databank is composed of simulations found from Zenodo repository with an appropriate licence. Most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

@@

The quality of each C--H bond order parameter is estimated by calculating the probability for a simulated value to locate within the error bars of the experimental value. Because conformational ensembles of individual lipids are independent in a fluid lipid bilayer, $\frac{S_{\rm CH}-\mu}{s/\sqrt{n}}$ has a Student's t-distribution with $n-1$ degrees of freedom, and the probability for an order parameter from simulation to locate within experimental error bars can be estimated from equation
\begin{equation}\label{probability}
  P = f \left( \frac{S_{\rm CH} - (S_{\rm exp}+\Delta S_{\rm exp})}{s/\sqrt{n}} \right) - f \left( \frac{S_{\rm CH} - (S_{\rm exp}-\Delta S_{\rm exp})}{s/\sqrt{n}} \right),
\end{equation}
where f(t) is the %first order 
Student's t-distribution, $\mu$ is the real mean of the order parameter, $n$ is the number of independent sample points for each C-H bond which equals the number of lipids in a simulation, $S_{\rm CH}$ is the sample mean from Eq.~\ref{OP}, $s$ is the variance of $S_{\rm CH}$ calculated over individual lipids, $S_{\rm exp}$ is the experimental value, and $\Delta S_{\rm exp}$ its error. The error of $\Delta S_{\rm exp} = 0.02$ is currently assumed for all experimental order parameters~\cite{ollila16}, yet more accurate may be available in the future~\cite{wurl22}. Because a lipid bilayer simulation contains at least dozens of lipids, the Student's t-distribution could be safely approximated with a normal distribution. However, the normal distribution gives probability values that are below the numerical accuracy of computers when simulation values are far from experiments. To avoid such numerical instabilities, we use the first order Student's t-distribution having slightly higher probabilities for values far away from the mean.
%Because phospholipids sample their conformational ensemble within nanosecond timescale~\cite{ferreira15}, all simulations in the databank would be sufficiently long to sample the realistic conformational phase of individual lipids. 
On the other hand, some force fields exhibit too slow dynamics which leads to large error bars in order parameter values~\cite{antila21a}. Such large error bars widen the Student's t-distribution in Eq.~\ref{probability} thereby artificially increasing the probability to find the simulated value within experimental error bars. Therefore, the order parameters with simulation error bars above the experimental error 0.02 are not included in the quality evaluation.

@@

Also a quality measure based on differences in simulated and experimental form factors accross available q-range is defined in the SIMtoEXP program~\cite{kucerka10}. However, the lobe heigths in simulated form factors depend on the simulation box size as shown in Fig.~\ref{fig:sizedependence}. Therefore, the quality measure defined in SIMtoEXP  would also depend on the simulation box size. Nevertheless, locations of form factor minima are independent on simulation box size in Fig.~\ref{fig:sizedependence}. Here we use the location of the first form factor minima for the quality evaluation because automatic detection of the location of second minima is inaccurate for some experimental data due to fluctuations, such as for the POPE data in Figs.~\ref{fig:quality} D and E. 
%In order to define a quality measure which is independent on simulation box size, we measure the form factor quality based on the location of the first minima observed in experiments. 
The first minimum correlates well with the thickness of a membrane (Fig.~\ref{fig:quality} F), although the correlation of the second minima would be even stronger (Fig.~\ref{fig:QualityCorrelationsSI}). In practise, we first filter the fluctuations from the form factor data using Savitzky-Golay filter (window lenght 30 and polynomial order 1) and locate the first minima above 0.1\,\AA$^{-1}$ from both simulation and experimental data. The quality of a form factor is then defined as Eucledian distance between the minima in simulated and experimental form factors, $FF_q = |FF_{\rm{min}}^{\rm{sim}}-FF_{\rm{min}}^{\rm{exp}}|$.

@@

\subsection{Calculation of lipid flip-flops}
Flip-flop rates were calculated using {\it AssignLeaflets} and {\it FlipFlop} tools from LiPyphilic package~\cite{LiPyphilic2021}. Headgroup atoms of each molecule as defined in the mapping file were used to determine in which leaflet the molecule locates. The midplane cut-off defining the region between leaflets was 1\,nm and frame cut-off was 100. This means that if the headgroup of a molecule entered within the distance of 1\,nm from the bilayer midplane and was found from the opposing leaflet after 100 steps, this event was considered as a successfull flip-flop event. The code that finds flip-flop events from all simulations in the NMRlipids databank is available at {\it scripts/FlipFlop.py} and the results at {\it Data/Flipflops/} in the repository at~\url{https://github.com/NMRLipids/DataBankManuscript/}.

\subsection{Analysing anisotropic diffusion of water in membrane environment from the NMRlipids databank}

@@

Water permeability through membranes was calculated from equation $P=r/2c_w$, where $r$ is the rate of permeation events per time and area, and $c_w$=33.3679\,(nm)$^{-3}$ is the concentration of water in bulk~\cite{venable19}. The number of permeation events in each trajectory was calculated using the code by Camilo et al.~\cite{camilo2022}, available at~\url{https://github.com/crobertocamilo/MD-permeation}. 
%This code was used within the loop that goes over all simulations in the NMRlipids databank and extracts the required information using the information in README.yaml files. 
The code that calculates permeabilities for all simulations in the NMRlipids databank is available at {\it /scripts/calcMD-PERMEATION.py} and the resulting permeabilities are stored at {\it /Data/MD-PERMEATION} in the repository containing all analyses specific for this publication at~\url{https://github.com/NMRLipids/DataBankManuscript/}. 
%While the order parameters, form factors, area per lipid and thickness are stored within the NMRlipids databank (\url{https://github.com/NMRLipids/Databank/tree/main/Data/Simulations}), further analyses can be conventiently stored in separate repositories with the same folder structure based on hash identities of trajectory and topology files. For example, results from further analyses performed here are stored in folders at \url{https://github.com/NMRLipids/DataBankManuscript/tree/main/Data}. 
This repository is organized similarly to the NMRlipids databank repository, enabling the upcycling of also the analyzed data without overloading the main NMRlipids databank repository.

@@
